---
title: "Untitled"
output: html_document
date: "2026-07-05"
---

### Rationale and selection of the bin width for the *fx* correction

The `fx` correction accounts for the frequency distribution of the target variable in the modern calibration dataset. This is relevant because tree-cover values are not represented uniformly: densely sampled portions of the gradient may have a disproportionate influence on model calibration, whereas uncommon values may be poorly represented. When `fx_method = "pspline"` is used, the frequency distribution is estimated and smoothed using a penalized spline. The `bin` parameter determines the width of the target intervals used to estimate this distribution and therefore controls the balance between resolution and stability. Very narrow bins preserve fine-scale variation but may produce noisy frequency estimates with limited observations per interval, whereas very wide bins provide more stable estimates but may obscure meaningful variation in the target gradient.

#Identification of pseudo-sites

To reduce the influence of spatial and environmental pseudoreplication during cross-validation, pseudo-sites were identified separately for each tree-cover target. Pseudo-sites are calibration samples that are both geographically close to the test sample and similar in their observed target value. In this analysis, similarity refers to tree cover rather than climate.

# Leave-out cross-validation

Predictive performance was evaluated separately for each region and tree-cover target using pseudo-site leave-out cross-validation. Unlike standard leave-one-out cross-validation, this procedure excludes not only the test sample but also calibration samples identified as its pseudo-sites. These samples are geographically close to the test site and have similar observed tree-cover values. Their exclusion reduces the risk of obtaining overly optimistic performance estimates due to spatial and environmental pseudoreplication.

For each cross-validation cycle, the fxTWA-PLS model was recalibrated using the remaining samples and then used to predict the tree-cover value of the excluded test sample. Models were evaluated with up to five PLS components, using the P-spline frequency correction and a target bin width of five percentage points. The same model settings were applied to all tree-cover targets to facilitate comparison among Gaussian weighting scenarios.

Cross-validation was conducted independently for each target because missing raster values resulted in slightly different sample compositions. Consequently, every target was evaluated using its corresponding pollen matrix, target vector, coordinate data, and pseudo-site structure. Results were saved separately immediately after completing each target to preserve completed calculations in the event of an interrupted run.

# Randomization t-test and selection of PLS components

A randomization t-test was applied to the cross-validation results to assess whether the addition of successive PLS components produced a statistically meaningful improvement in predictive performance. The test was conducted independently for every region and tree-cover target using 999 permutations.

The procedure compares the reduction in cross-validated prediction error associated with each additional component against a distribution generated through randomization. Components were therefore assessed using both the change in root mean squared error of prediction (ΔRMSEP) and its associated permutation-based probability. A low probability indicates that the observed improvement in predictive performance is unlikely to have occurred by chance.

The number of components should not be selected solely from the model with the lowest calibration error, as calibration error generally decreases when model complexity increases. Instead, component selection was based on cross-validated performance and the randomization test. Preference was given to the smallest model that captured the reproducible predictive signal, avoiding additional components that did not provide a statistically supported reduction in RMSEP.

Randomization-test results were combined into a single table containing the region, tree-cover target, component number, cross-validation statistics, and permutation-test results. This allowed component performance to be compared consistently among Gaussian weighting scenarios and between Europe and Latin America.



```{r}
#You can install the released version of fxTWAPLS from CRAN with:
#install.packages("fxTWAPLS")
library(tidyverse)
library(fxTWAPLS)
library(foreach)
```


```{r}
# Load and prepare land polygon for South America
land_polyg_latam <- sf::st_read(
  "H:/Mi unidad/Research/Natural_earth_data/Land_polygon_large_ne_10m_land/ne_10m_land.shp"
) |>
  sf::st_transform(crs = 4326) |>
  sf::as_Spatial() |>
  raster::crop(raster::extent(-118, -33, -55, 32)) |>
  sf::st_as_sf()

ggplot() + geom_sf(data = land_polyg_latam)


# Load and prepare land polygon for Europe
land_polyg_eur <- sf::st_read(
  "H:/Mi unidad/Research/Natural_earth_data/Land_polygon_large_ne_10m_land/ne_10m_land.shp") |>
  sf::st_transform(crs = 4326) |>
  sf::as_Spatial() |>
  raster::crop(raster::extent(-20, 62, 29, 75)) |> #min,x, max.x, min.y, max.y
  sf::st_as_sf()

ggplot() + geom_sf(data = land_polyg_eur)
```


1. Setup – define constants and load data
```{r}
iteration <- "i3" 
my_path <-"Data/Processed/Output_calibration_and_reconstruction/"

```




```{r}
# Load data (as before)
europe_extraction <- readRDS("Data/Processed/tree_gaussian_EUROPA_25_50km_modern_pollen.rds") |>
  select(id_sample:hill_n2, starts_with("Tree_"), everything())
#set.seed(9751); europe_extraction <- sample_n(europe_extraction, size = 250)

latam_extraction <- readRDS("Data/Processed/tree_gaussian_LATAM_25_50km_modern_pollen.rds") |>
  select(id_sample:hill_n2, starts_with("Tree_"), everything())
#set.seed(9751); latam_extraction <- sample_n(latam_extraction, size = 250)
```







Chunk 1: Initial set up and load data
```{r}

# Definir las columnas target (las de extracción de raster)
target_cols <- c( "Tree_Gaussian_50_sigma100", "Tree_Gaussian_50_sigma10",  "Tree_Gaussian_50_sigma3",   "Tree_Gaussian_50_sigma2",  "Tree_Gaussian_25_sigma100", "Tree_Gaussian_25_sigma10" , "Tree_Gaussian_25_sigma3" ,  "Tree_Gaussian_25_sigma2" )


# Identificar las columnas que NO son target (identificadores + taxa)
# Para LATAM
id_cols <- c("id_sample" ,    "latitude" ,        "longitude"    ,    "elevation"         ,     "hill_n2" )
```




Identificar los taxones de cada dataset
Y crear un da
```{r}
#=========== LATAM ========
taxa_cols_latam <- setdiff(names(latam_extraction), 
                          c(id_cols, target_cols))

# Crear una lista de dataframes para LATAM, cada uno con un target diferente
latam_target_list <- list()

for (target in target_cols) {
  # Seleccionar: identificadores + el target actual + todas las taxa
  latam_target_list[[target]] <- latam_extraction %>%
    select(all_of(c(id_cols, target, taxa_cols_latam)))
}



#======= EUROPE ========
taxa_cols_europe <- setdiff(names(europe_extraction), 
                          c(id_cols, target_cols))

# Crear una lista de dataframes para EUROPE, cada uno con un target diferente
europe_target_list <- list()

for (target in target_cols) {
  # Seleccionar: identificadores + el target actual + todas las taxa
  europe_target_list[[target]] <- europe_extraction %>%
    select(all_of(c(id_cols, target, taxa_cols_europe)))
}




latam_target_list[["Tree_Gaussian_25_sigma2"]]
europe_target_list[["Tree_Gaussian_25_sigma2"]]
```


Puedes filtrar cada dataframe según su propio target usando imap() de purrr. El nombre de cada elemento de la lista coincide con el nombre de su target.
```{r}
latam_target_list <- latam_target_list |>
  purrr::imap(\(df, target) 
    df |>
      filter(!is.na(.data[[target]]))
  )

europe_target_list <- europe_target_list |>
  purrr::imap(\(df, target) 
    df |>
      filter(!is.na(.data[[target]]))
  )

latam_target_list[["Tree_Gaussian_25_sigma2"]]
europe_target_list[["Tree_Gaussian_25_sigma2"]]
```

Chunk 1: Extraer componentes para LATAM
Está desarmando cada dataframe en las cuatro piezas necesarias para aplicar posteriormente TWAPLS.
latam_twapls_data will have a structure like this: 

latam_twapls_data
├── Tree_Gaussian_50_sigma100
│   ├── taxa
│   ├── target
│   ├── coords
│   └── ids
├── Tree_Gaussian_50_sigma10
│   ├── taxa
│   ├── target
│   ├── coords
│   └── ids
...
└── Tree_Gaussian_25_sigma2
    ├── taxa
    ├── target
    ├── coords
    └── ids
```{r}
# Extraer componentes para LATAM
latam_twapls_data <- map2(
  latam_target_list,
  names(latam_target_list),
  ~ {
    df <- .x
    target <- .y
    
    # Busca cuáles columnas de id_cols realmente existen en el dataframe.
    id_cols_present <- intersect(id_cols, names(df))
    
    # El target y las columnas fijas se excluyen de las taxa
    exclude <- c(id_cols_present, target)
    taxa_cols <- setdiff(names(df), exclude)
    
    list(
      taxa = df %>% select(all_of(taxa_cols)),
      target = df[[target]],
      coords = df %>% select(longitude, latitude),
      ids = df$id_sample
    )
  }
)

latam_twapls_data[["Tree_Gaussian_25_sigma2"]]$target
```


```{r}
saveRDS(latam_twapls_data, "Data/Processed/Output_calibration_and_reconstruction/latam_twapls_data.rds")

```


```{r}
# Extraer componentes para EUROPA
europe_twapls_data <- purrr::map2(
  europe_target_list,
  names(europe_target_list),
  ~ {
    df <- .x
    target <- .y
    
    # Identificar las columnas fijas presentes
    id_cols_present <- intersect(id_cols, names(df))
    
    # Excluir identificadores y el target de las columnas de taxones
    exclude <- c(id_cols_present, target)
    taxa_cols <- setdiff(names(df), exclude)
    
    list(
      taxa = df |> select(all_of(taxa_cols)),
      target = df[[target]],
      coords = df |> select(longitude, latitude),
      ids = df$id_sample
    )
  }
)

europe_twapls_data[["Tree_Gaussian_25_sigma2"]]$taxa
```


```{r}
saveRDS(europe_twapls_data, "Data/Processed/Output_calibration_and_reconstruction/europe_twapls_data.rds")

```



In the European dataset, the target distribution was strongly concentrated at low tree-cover values, followed by lower and irregular frequencies across the remaining range, although no intervals were empty. Bin widths of 2.5, 5, and 10 percentage points were compared. With bin = 2.5, the first interval contained 43 samples (17.8% of the dataset), while the remaining intervals had a median of six samples. With bin = 5, the first interval contained 49 samples (20.3%) and the remaining intervals had a median of 11 samples. With bin = 10, the corresponding values were 63 samples (26.1%) and a median of 21 samples.

A bin width of five percentage points was selected as a practical compromise. It reduced small-scale irregularity and provided greater support per interval than bin = 2.5, while retaining more information across the 0–100% tree-cover gradient than bin = 10. The high frequency in the first interval was retained because it represents a genuine feature of the calibration dataset rather than missing or erroneous data. Nevertheless, the final choice of bin width should be supported by cross-validation and sensitivity analyses to confirm that model performance and the selected number of PLS components are stable across reasonable alternative bin widths.
```{r}
target_x <- europe_twapls_data[["Tree_Gaussian_25_sigma2"]]$target


purrr::map_dfr(c(2.5, 5, 10), \(bin_width) {
    
    breaks_x <- seq(0, ceiling(max(target_x) / bin_width) * bin_width, by = bin_width)
    
    counts <- hist(target_x, breaks = breaks_x, plot = FALSE, include.lowest = TRUE)$counts
    
    tibble(bin = bin_width,first_bin_n = counts[1],
      first_bin_percent = counts[1] / sum(counts) * 100,
      median_other_bins = median(counts[-1]),
      maximum_other_bin = max(counts[-1])
    )
  }
)

target_df <- tibble(target = target_x)

p2 <- ggplot(target_df, aes(target)) +
  geom_histogram(binwidth = 2, boundary = 0) +
  theme_classic() +
  labs(title = "bin = 2.")

p2.5 <- ggplot(target_df, aes(target)) +
  geom_histogram(binwidth = 2.5, boundary = 0) +
  theme_classic() +
  labs(title = "bin = 2.5")

p5 <- ggplot(target_df, aes(target)) +
  geom_histogram(binwidth = 5, boundary = 0) +
  theme_classic() +
  labs(title = "bin = 5")

p10 <- ggplot(target_df, aes(target)) +
  geom_histogram(binwidth = 10, boundary = 0) +
  theme_classic() +
  labs(title = "bin = 10")

p2
p2.5
p5
p10
```


```{r}
target_x <- latam_twapls_data[["Tree_Gaussian_25_sigma2"]]$target


purrr::map_dfr(c(2.5, 5, 10), \(bin_width) {
    
    breaks_x <- seq(0, ceiling(max(target_x) / bin_width) * bin_width, by = bin_width)
    
    counts <- hist(target_x, breaks = breaks_x, plot = FALSE, include.lowest = TRUE)$counts
    
    tibble(bin = bin_width,first_bin_n = counts[1],
      first_bin_percent = counts[1] / sum(counts) * 100,
      median_other_bins = median(counts[-1]),
      maximum_other_bin = max(counts[-1])
    )
  }
)

target_df <- tibble(target = target_x)

p2 <- ggplot(target_df, aes(target)) +
  geom_histogram(binwidth = 2, boundary = 0) +
  theme_classic() +
  labs(title = "bin = 2.")

p2.5 <- ggplot(target_df, aes(target)) +
  geom_histogram(binwidth = 2.5, boundary = 0) +
  theme_classic() +
  labs(title = "bin = 2.5")

p5 <- ggplot(target_df, aes(target)) +
  geom_histogram(binwidth = 5, boundary = 0) +
  theme_classic() +
  labs(title = "bin = 5")

p10 <- ggplot(target_df, aes(target)) +
  geom_histogram(binwidth = 10, boundary = 0) +
  theme_classic() +
  labs(title = "bin = 10")

p2
p2.5
p5
p10
```



```{r, warning=FALSE}
fit_targets_europe <- europe_twapls_data |>
  purrr::map(
    \(data) {
      fxTWAPLS::TWAPLS.w2(
        data$taxa,
        data$target,
        nPLS = 5,
        usefx = TRUE,
        fx_method = "pspline",
        bin = 2
      )
    }
  )



fit_targets_latam <- latam_twapls_data |>
  purrr::map(
    \(data) {
      fxTWAPLS::TWAPLS.w2(
        data$taxa,
        data$target,
        nPLS = 5,
        usefx = TRUE,
        fx_method = "pspline",
        bin = 2.5
      )
    }
  )
```


```{r}
saveRDS(fit_targets_europe,
  "Data/Processed/Output_calibration_and_reconstruction/fit_targets_europe_bin2.rds")

saveRDS(fit_targets_latam,
  "Data/Processed/Output_calibration_and_reconstruction/fit_targets_latam_bin2.5.rds")
```



Because samples with missing target values were removed independently for each target, the resulting datasets differed slightly in their sample composition. Geographic distance matrices were therefore calculated separately for every combination of region and tree-cover target, ensuring that the rows and columns of each matrix matched the corresponding pollen, target, coordinate, and sample-ID data.

Pairwise geographic distances were calculated from sample longitude and latitude using get_distance(). The resulting distance matrix and the corresponding tree-cover target were then supplied to get_pseudo(). During cross-validation, the test sample and its associated pseudo-sites can be excluded from model training, providing a more conservative assessment of predictive performance by reducing the contribution of geographically close samples with similar tree-cover conditions.
```{r}
CPUS <- 8

#Distancias para Eruopa
dist_targets_europe <- europe_twapls_data |>
  purrr::imap(
    \(data, target_name) {
      
      message("Calculating distances for: ", target_name)
      
      fxTWAPLS::get_distance(
        data$coords,
        cpus = CPUS
      )
    }
  )


#Pseudosites para Europa
pseudo_targets_europe <- purrr::map2(
  dist_targets_europe,
  europe_twapls_data,
  \(distance_matrix, data) {
    
    fxTWAPLS::get_pseudo(
      distance_matrix,
      data$target,
      cpus = CPUS
    )
  }
)
```


```{r}
saveRDS(dist_targets_europe, "Data/Processed/Output_calibration_and_reconstruction/dist_targets_europe.rds")


saveRDS(pseudo_targets_europe, "Data/Processed/Output_calibration_and_reconstruction/pseudo_targets_europe.rds")

```




```{r}
#Distancias para LATAM
dist_targets_latam <- latam_twapls_data |>
  purrr::imap(
    \(data, target_name) {
      
      message("Calculating distances for: ", target_name)
      
      fxTWAPLS::get_distance(
        data$coords,
        cpus = CPUS
      )
    }
  )


#Pseudosites para LATAM
pseudo_targets_latam <- purrr::map2(
  dist_targets_latam,
  latam_twapls_data,
  \(distance_matrix, data) {
    
    fxTWAPLS::get_pseudo(
      distance_matrix,
      data$target,
      cpus = CPUS
    )
  }
)
```

```{r}
saveRDS(dist_targets_latam, "Data/Processed/Output_calibration_and_reconstruction/dist_targets_latam.rds")

saveRDS(pseudo_targets_latam, "Data/Processed/Output_calibration_and_reconstruction/pseudo_targets_latam.rds")
```


```{r}
CPUS <- 9
iteration <- "i3"

cv_path <- "Data/Processed/Output_calibration_and_reconstruction/Cross_validation/"

dir.create(
  cv_path,
  recursive = TRUE,
  showWarnings = FALSE
)
```


Validacion cruzada para Europa


```{r}
cv_targets_europe <- purrr::imap(
  europe_twapls_data,
  \(data, target_name) {
    
    message("Cross-validating Europe: ", target_name)
    
    pseudo <- pseudo_targets_europe[[target_name]]
    
    result <- fxTWAPLS::cv.pr.w(
      data$taxa,
      data$target,
      nPLS = 5,
      fxTWAPLS::TWAPLS.w2,
      fxTWAPLS::TWAPLS.predict.w,
      pseudo,
      usefx = TRUE,
      fx_method = "pspline",
      bin = 2,
      cpus = CPUS,
      test_mode = FALSE
    ) |>
      fxTWAPLS::pb()
    
    write.csv(
      result,
      file.path(
        cv_path,
        paste0(
          "cv_europe_",
          target_name,
          "_bin5_",
          iteration,
          ".csv"
        )
      ),
      row.names = FALSE
    )
    
    result
  }
)
```

Validacion cruzada para LATAM

```{r}
pseudo_targets_latam <- readRDS("Data/Processed/Output_calibration_and_reconstruction/pseudo_targets_latam.rds")

latam_twapls_data <- readRDS("Data/Processed/Output_calibration_and_reconstruction/latam_twapls_data.rds")
```


```{r, warning=FALSE}
cv_targets_latam <- purrr::imap(
  latam_twapls_data,
  \(data, target_name) {
    
    message("Cross-validating LATAM: ", target_name)
    
    pseudo <- pseudo_targets_latam[[target_name]]
    
    result <- fxTWAPLS::cv.pr.w(
      data$taxa,
      data$target,
      nPLS = 5,
      fxTWAPLS::TWAPLS.w2,
      fxTWAPLS::TWAPLS.predict.w,
      pseudo,
      usefx = TRUE,
      fx_method = "pspline",
      bin = 2.5,
      cpus = CPUS,
      test_mode = FALSE
    ) |>
      fxTWAPLS::pb()
    
    write.csv(
      result,
      file.path(
        cv_path,
        paste0(
          "cv_latam_",
          target_name,
          "_bin5_",
          iteration,
          ".csv"
        )
      ),
      row.names = FALSE
    )
    
    result
  }
)
```

Cargar los archivos individuales 
```{r}
cv_path <- "C:/Research/Tree_cover_LATAM_EMBSECBIO/Data/Processed/Output_calibration_and_reconstruction/Cross_validation/europe_bin5/"
bin = 5

cv_files_europe <- list.files(
  cv_path,
  pattern = "^cv_europe_Tree_Gaussian_.*_bin5_i3.1\\.csv$",
  full.names = TRUE)

basename(cv_files_europe)
length(cv_files_europe)



target_names_europe <- basename(cv_files_europe) |>
  stringr::str_remove("^cv_europe_") |>
  stringr::str_remove("_bin5\\.5_i3.1\\.csv$")


#4. Reconstruir la lista cv_targets europe
cv_targets_europe <- cv_files_europe |>
  purrr::map(readr::read_csv, show_col_types = FALSE) |> rlang::set_names(target_names_europe)


cv_targets_europe <- cv_targets_europe |>
  purrr::map(as.data.frame)
```


```{r}
cv_path <- "C:/Research/Tree_cover_LATAM_EMBSECBIO/Data/Processed/Output_calibration_and_reconstruction/Cross_validation/latam_bin3.5/"
bin = 5

cv_files_latam <- list.files(
  cv_path,
  pattern = "^cv_latam_Tree_Gaussian_.*_bin3.5_i3.1\\.csv$",
  full.names = TRUE)

basename(cv_files_latam)
length(cv_files_latam)



target_names_latam <- basename(cv_files_latam) |>
  stringr::str_remove("^cv_europe_") |>
  stringr::str_remove("_bin3.5\\.5_i3.1\\.csv$")


#4. Reconstruir la lista cv_targets europe
cv_targets_latam <- cv_files_latam |>
  purrr::map(readr::read_csv, show_col_types = FALSE) |> rlang::set_names(target_names_latam)


cv_targets_latam <- cv_targets_latam |>
  purrr::map(as.data.frame)
```



Random t-test 
```{r}
#para Europa
set.seed(9751)

rand_targets_europe <- cv_targets_europe |>
  purrr::imap(
    \(cv_result, target_name) {
      
      message("Random t-test for Europe: ", target_name)
      
      fxTWAPLS::rand.t.test.w(
        cv_result,
        n.perm = 999
      ) |>
        as_tibble() |>
        mutate(
          region = "Europe",
          target = target_name,
          n_comp = row_number(),
          .before = 1
        )
    }
  )

#para LATAM
set.seed(9751)

rand_targets_latam <- cv_targets_latam |>
  purrr::imap(
    \(cv_result, target_name) {
      
      message("Random t-test for LATAM: ", target_name)
      
      fxTWAPLS::rand.t.test.w(
        cv_result,
        n.perm = 999
      ) |>
        as_tibble() |>
        mutate(
          region = "LATAM",
          target = target_name,
          n_comp = row_number(),
          .before = 1
        )
    }
  )
```


```{r}
rand_table_europe <- bind_rows(rand_targets_europe)

rand_table_latam <- bind_rows(rand_targets_latam)
```


```{r}
rand_table_all <- bind_rows(
  rand_table_europe,
  rand_table_latam) |>
  select(region, target, n_comp, everything())
```


```{r}
write.csv(rand_table_europe, "Data/Processed/Output_calibration_and_reconstruction/Cross_validation/random_t_test_europe_bin5.csv", row.names = FALSE)

write.csv(rand_table_latam, "Data/Processed/Output_calibration_and_reconstruction/Cross_validation/random_t_test_latam_bin3.5.csv", row.names = FALSE)

```


```{r}
write.csv(
  rand_table_all,
  "Data/Processed/Output_calibration_and_reconstruction/Cross_validation/random_t_test_all_regions_bin3.5latam_and_5europe.csv",
  row.names = FALSE)
```


```{r}
selected_components <- rand_table_all |>
  group_by(region, target) |>
  summarise(
    selected_n_comp = if_else(
      any(p < 0.05, na.rm = TRUE),
      max(n_comp[p < 0.05], na.rm = TRUE),
      NA_integer_
    ),
    .groups = "drop"
  )
```


```{r}
write.csv(
  selected_components,
  file.path(
    cv_path,
    paste0(
      "selected_components_all_regions_bin5_",
      iteration,
      ".csv"
    )
  ),
  row.names = FALSE
)
```

#Despues de una seleccion manual de los componentes

```{r}
plot_data_europe <- cv_targets_europe[["Tree_Gaussian_50_sigma3"]] |>
  as_tibble() |>
  transmute(
    region = "Europe",
    target = "Tree_Gaussian_50_sigma3",
    component = 3L,
    id_sample = europe_twapls_data[["Tree_Gaussian_50_sigma3"]]$ids,
    observed = test.x,
    predicted = comp3
  )


plot_data_latam <- cv_targets_latam[["Tree_Gaussian_25_sigma3"]] |>
  as_tibble() |>
  transmute(
    region = "LATAM",
    target = "Tree_Gaussian_25_sigma3",
    component = 2L,
    id_sample = latam_twapls_data[["Tree_Gaussian_25_sigma3"]]$ids,
    observed = test.x,
    predicted = comp2
  )

plot_data_final <- bind_rows(
  plot_data_europe,
  plot_data_latam
)

glimpse(plot_data_final)

plot_data_final |>
  count(region)
```






##################################################
#####################################
#########################
##################
```{r}
#=========== LATAM ========
taxa_cols_latam <- setdiff(names(latam_extraction), c(id_cols, target_cols_latam))
# Crear una lista de dataframes para LATAM, cada uno con un target diferente
latam_target_list <- list()
for (target in target_cols_latam) {
  # Seleccionar: identificadores + el target actual + todas las taxa
  latam_target_list[[target]] <- latam_extraction %>%
    select(all_of(c(id_cols, target, taxa_cols_latam)))
}



#======= EUROPE ========
taxa_cols_europe <- setdiff(names(europe_extraction), c(id_cols, target_cols_europe))

europe_target_list <- list()
for (target in target_cols_europe) {
  # Seleccionar: identificadores + el target actual + todas las taxa
  europe_target_list[[target]] <- europe_extraction %>%
    select(all_of(c(id_cols, target, taxa_cols_europe)))
}



# Calcular varianza y frecuencia
nzv_metrics <- nearZeroVar(europe_extraction[, taxa_cols_europe], 
                           freqCut = 95/5,  # Valor más frecuente >95%, segundo <5%
                           uniqueCut = 10,   # Número de valores únicos
                           saveMetrics = TRUE)

# Ver cuántos taxones se van a eliminar
cat("Taxones totales:", length(taxa_cols_europe), "\n")
cat("Taxones a eliminar:", sum(nzv_metrics$nzv), "\n")

# Eliminar taxones con varianza cercana a cero
taxa_cols_europe <- setdiff(taxa_cols_europe, 
                            rownames(nzv_metrics)[nzv_metrics$nzv == TRUE])

cat("Taxones que se mantienen:", length(taxa_cols_europe), "\n")

```




Chunk 1: Extraer componentes para LATAM
```{r}
# Correct way to create the list for LATAM
latam_twapls_data <- list()

for (target in target_cols_latam) {
  # Select columns for this target
  df_temp <- latam_extraction %>%
    select(all_of(c(id_cols, target, taxa_cols_latam)))
  
  latam_twapls_data[[target]] <- list(
    taxa = df_temp %>% select(all_of(taxa_cols_latam)),
    target = df_temp[[target]],
    coords = df_temp %>% select(longitude, latitude),
    ids = df_temp$id_sample
  )
}

# Similarly for EUROPE
europe_twapls_data <- list()
for (target in target_cols_europe) {
  df_temp <- europe_extraction %>%
    select(all_of(c(id_cols, target, taxa_cols_europe)))
  
  europe_twapls_data[[target]] <- list(
    taxa = df_temp %>% select(all_of(taxa_cols_europe)),
    target = df_temp[[target]],
    coords = df_temp %>% select(longitude, latitude),
    ids = df_temp$id_sample
  )
}
```




Chunk 2: Configurar paralelización y ejecutar TWAPLS para LATAM


```{r, warning=FALSE}
# Set sequential processing globally
library(future)
plan(sequential)

total_cores <- parallel::detectCores()
# Opción 3: Usar 7 (la mitad, pero podría tener problemas)
CPUS <- floor(total_cores / 2)  7



# Ejecutar TWAPLS para LATAM (todos los targets)
latam_results <- map2(
  latam_twapls_data,
  names(latam_twapls_data),
  ~ {
    cat("\n=== LATAM -", .y, "===\n")
    
    taxa <- .x$taxa
    target <- .x$target
    coords <- .x$coords
    
    # 1. Ajuste del modelo
    fit <- TWAPLS.w2(taxa, target, nPLS = 5, 
                     usefx = TRUE, fx_method = "pspline", bin = 0.08)
    
    # 2. Matriz de distancias
    dist_mat <- get_distance(coords, cpus = CPUS)
    
    # 3. Pseudo‑sitios para validación cruzada
    pseudo <- get_pseudo(dist_mat, target, cpus = CPUS)
    
    # 4. Validación cruzada - USANDO la sintaxis de tu código original
    cv_res <- cv.pr.w(taxa, target, nPLS = 5,
                      TWAPLS.w2,           # fit.fun (posicional)
                      TWAPLS.predict.w,    # pred.fun (posicional)
                      pseudo = pseudo,
                      usefx = TRUE,
                      fx_method = "pspline",
                      bin = 0.08,
                      cpus = CPUS,
                      test_mode = FALSE) %>% pb()
    
    # 5. Test aleatorio de permutaciones
    rand_test <- rand.t.test.w(cv_res, n.perm = 999) %>%
      as_tibble() %>%
      mutate(comp = row_number())
    
    # 6. Mejor componente
    best_comp <- rand_test %>%
      filter(p < 0.01) %>%
      slice_min(RMSEP) %>%
      pull(comp)
    
     # 6. Mejor componente
    p_val <- rand_test %>%
      filter(p < 0.01) %>%
      slice_min(RMSEP) %>%
      pull(p)
    
    if (length(best_comp) == 0) {
      best_comp <- rand_test %>% slice_min(RMSEP) %>% pull(comp)
      cat("  -> Ningún componente significativo, se usa el de menor RMSEP (comp", best_comp, ")\n")
    } else {
      cat("  -> Mejor componente:", best_comp, "\n")
    }
    
    # 7. Observados vs predichos
    obs_pred <- data.frame(
      Observed = cv_res$test.x,
      Predicted = cv_res[[paste0("comp", best_comp)]]
    )
    
    # 8. Métricas
    rmsep <- sqrt(mean((obs_pred$Observed - obs_pred$Predicted)^2))
    r2 <- cor(obs_pred$Observed, obs_pred$Predicted)^2
    
    performance <- tibble(
      region = "LATAM",
      target = .y,
      best_comp = best_comp,
      p_val = p_val,
      RMSEP = rmsep,
      R2 = r2
    )
    
    list(
      target = .y,
      region = "LATAM",
      cv = cv_res,
      rand_test = rand_test,
      best_comp = best_comp,
      obs_pred = obs_pred,
      performance = performance
    )
  }
)

```


```{r}
saveRDS(latam_results, file = paste(my_path,"latam_tree_recons_results_",iteration,".rds", sep=""))
```

## EUROPE

```{r}
# Set sequential processing globally
library(future)
plan(sequential)

total_cores <- parallel::detectCores()

CPUS <- 8


# ============================================================
# MATRIZ DE DISTANCIAS DE EUROPE
# ============================================================
# Las coordenadas son las mismas para todos los escenarios,
# por lo que calculamos las distancias UNA SOLA VEZ.

coords_europe <- europe_twapls_data[[1]]$coords

dist_mat_europe <- get_distance(
  coords_europe,
  cpus = CPUS
)

cat("\n=== Matriz de distancias de EUROPE calculada ===\n")


saveRDS(
  dist_mat_europe,
  file = paste(
    my_path,
    "europe_distance_matrix_",
    iteration,
    ".rds",
    sep = ""
  )
)



dist_mat_europe <- readRDS(file = paste(
    my_path,
    "europe_distance_matrix_",
    iteration,
    ".rds",
    sep = ""
  ))
```


```{r, warning=FALSE}
# ============================================================
# EUROPE: UN ESCENARIO A LA VEZ
# ============================================================

for (target_name in names(europe_twapls_data)) {
  
  cat("\n")
  cat("========================================\n")
  cat("=== EUROPE -", target_name, "===\n")
  cat("========================================\n")
  
  
  # ----------------------------------------------------------
  # Archivo donde se guardará este escenario
  # ----------------------------------------------------------
  
  output_file <- paste(
    my_path,
    "europe_",
    target_name,
    "_",
    iteration,
    ".rds",
    sep = ""
  )
  
  
  # ----------------------------------------------------------
  # Si ya existe, no recalcular
  # ----------------------------------------------------------
  
  if (file.exists(output_file)) {
    
    cat(
      "-> Este escenario ya existe.\n",
      "-> Se salta y continúa con el siguiente.\n"
    )
    
    next
  }
  
  
  # ----------------------------------------------------------
  # Extraer datos del escenario
  # ----------------------------------------------------------
  
  data_current <- europe_twapls_data[[target_name]]
  
  taxa <- data_current$taxa
  target <- data_current$target
  coords <- data_current$coords
  
  
  # ----------------------------------------------------------
  # 1. Ajuste del modelo
  # ----------------------------------------------------------
  
  cat("1. Ajustando TWAPLS...\n")
  
  fit <- TWAPLS.w2(
    taxa,
    target,
    nPLS = 5,
    usefx = TRUE,
    fx_method = "pspline",
    bin = 2.5
  )
  
  
  # ----------------------------------------------------------
  # 2. Pseudo-sitios
  #
  # IMPORTANTE:
  # Se calcula nuevamente para cada escenario
  # porque depende de 'target'
  # ----------------------------------------------------------
  
  cat("2. Calculando pseudo-sitios...\n")
  
  pseudo <- get_pseudo(
    dist_mat_europe,
    target,
    cpus = CPUS
  )
  
  
  # ----------------------------------------------------------
  # 3. Validación cruzada
  # ----------------------------------------------------------
  
  cat("3. Ejecutando validación cruzada...\n")
  
  cv_res <- cv.pr.w(
    taxa,
    target,
    nPLS = 5,
    TWAPLS.w2,
    TWAPLS.predict.w,
    pseudo = pseudo,
    usefx = TRUE,
    fx_method = "pspline",
    bin = 2.5,
    cpus = CPUS,
    test_mode = FALSE
  ) %>% pb()
  
  
  # ----------------------------------------------------------
  # 4. Test aleatorio
  # ----------------------------------------------------------
  
  cat("4. Ejecutando randomisation test...\n")
  
  rand_test <- rand.t.test.w(
    cv_res,
    n.perm = 999
  ) %>%
    as_tibble() %>%
    mutate(comp = row_number())
  
  
  # ----------------------------------------------------------
  # 5. Mejor componente
  # ----------------------------------------------------------
  
  best_comp <- rand_test %>%
    filter(p < 0.01) %>%
    slice_min(RMSEP) %>%
    pull(comp)
  
  
  p_val <- rand_test %>%
    filter(p < 0.01) %>%
    slice_min(RMSEP) %>%
    pull(p)
  
  
  if (length(best_comp) == 0) {
    
    best_comp <- rand_test %>%
      slice_min(RMSEP) %>%
      pull(comp)
    
    cat(
      "  -> Ningún componente significativo,",
      "se usa el de menor RMSEP (comp", 
      best_comp, ")\n"
    )
    
  } else {
    
    cat(
      "  -> Mejor componente:",
      best_comp,
      "\n"
    )
  }
  
  
  # ----------------------------------------------------------
  # 6. Observados vs predichos
  # ----------------------------------------------------------
  
  obs_pred <- data.frame(
    Observed = cv_res$test.x,
    Predicted = cv_res[[paste0("comp", best_comp)]]
  )
  
  
  # ----------------------------------------------------------
  # 7. Métricas
  # ----------------------------------------------------------
  
  rmsep <- sqrt(
    mean(
      (obs_pred$Observed - obs_pred$Predicted)^2
    )
  )
  
  r2 <- cor(
    obs_pred$Observed,
    obs_pred$Predicted
  )^2
  
  
  performance <- tibble(
    region = "EUROPE",
    target = target_name,
    best_comp = best_comp,
    p_val = p_val,
    RMSEP = rmsep,
    R2 = r2
  )
  
  
  # ----------------------------------------------------------
  # 8. Crear resultado
  # ----------------------------------------------------------
  
  result <- list(
    target = target_name,
    region = "EUROPE",
    cv = cv_res,
    rand_test = rand_test,
    best_comp = best_comp,
    obs_pred = obs_pred,
    performance = performance
  )
  
  
  # ----------------------------------------------------------
  # 9. GUARDAR INMEDIATAMENTE
  # ----------------------------------------------------------
  
  saveRDS(
    result,
    output_file
  )
  
  cat(
    ">>> ESCENARIO GUARDADO:",
    target_name,
    "\n"
  )
  
  cat(
    ">>> Archivo:",
    output_file,
    "\n"
  )
}
```



```{r}
# ============================================================
# RECONSTRUIR EUROPE_RESULTS
# ============================================================

europe_files <- list.files(
  path = my_path,
  pattern = paste0(
    "^europe_EUROPA.*_",
    iteration,
    "\\.rds$"
  ),
  full.names = TRUE
)


europe_results <- lapply(
  europe_files,
  readRDS
)


names(europe_results) <- sapply(
  europe_results,
  function(x) x$target
)


europe_performance <- purrr::map_dfr(
  europe_results,
  ~ .x$performance
)

europe_performance


europe_rand <- purrr::map_dfr(
  europe_results,
  ~ .x$rand_test |> 
    mutate(target = .x$target, .before = 1)
)

europe_rand
```



```{r}
write.csv(europe_rand, file = "C:/Research/Tree_cover_LATAM_EMBSECBIO/Data/Processed/Output_calibration_and_reconstruction/europe_rand_test_i3.csv")
```













```{r}
saveRDS(europe_results, file = paste(my_path,"europe_tree_recons_results_",iteration,".rds", sep=""))
```



```{r}
latam_results <- readRDS("Data/Processed/Output_calibration_and_reconstruction/latam_tree_recons_results_i3.rds")
```






Chunk 1: Identificar el mejor target
```{r}
# LATAM - identificar mejor target (menor RMSEP)
latam_performance <- map_dfr(latam_results, ~ .x$performance)

latam_random_test <- map_dfr(latam_results, ~ .x$rand_test)

best_latam_target <- latam_performance %>% slice_min(RMSEP) %>% pull(target)

#best_latam_rmsep <- latam_performance %>%
  #slice_min(RMSEP) %>%
  #pull(RMSEP)

latam_performance

cat("Mejor target en LATAM:", best_latam_target)
#--------

# LATAM - identificar mejor target (menor RMSEP)
europe_performance <- map_dfr(europe_results, ~ .x$performance)

best_europe_target <- europe_performance %>% slice_min(RMSEP) %>% pull(target)

#best_latam_rmsep <- latam_performance %>%
  #slice_min(RMSEP) %>%
  #pull(RMSEP)

europe_performance

cat("Mejor target en EUROPE:", best_europe_target)
```




```{r}
write.csv(latam_random_test, 
          file = "C:/Research/Tree_cover_LATAM_EMBSECBIO/Data/Processed/Output_calibration_and_reconstruction/latam_rand_test_i3.csv")
```


```{r}
write.csv(latam_performance, "Data/Processed/Output_calibration_and_reconstruction/latam_performance_i3.csv")
```


```{r}
write.csv(europe_performance, "Data/Processed/Output_calibration_and_reconstruction/europe_performance_i3.csv")
```


Chunk 2: Crear dataframe diagnóstico para LATAM (mejor target)
```{r}
# Extraer el mejor resultado de LATAM
latam_best_result <- latam_results[[best_latam_target]]

# Extraer los datos preparados originales para ese target
latam_best_data <- latam_extraction[[best_latam_target]]


# Verificar que tienen el mismo número de filas
cat("Número de filas en obs_pred:", nrow(latam_best_result$obs_pred), "\n")
cat("Número de filas en datos preparados:", nrow(latam_best_data), "\n")



# Crear dataframe diagnóstico para LATAM
latam_diagnostic <- latam_best_data %>%
  # Unir con observed y predicted
  bind_cols(
    Observed = latam_best_result$obs_pred$Observed,
    Predicted = latam_best_result$obs_pred$Predicted) %>%
  # Calcular métricas de error
  mutate(
    Residual = Observed - Predicted,
    Abs_Residual = abs(Residual),
    Pct_Error = (Abs_Residual / (Observed + 0.001)) * 100,  # +0.001 para evitar división por cero
    Squared_Error = (Observed - Predicted)^2,
    Relative_Error = ifelse(Observed > 0, (Observed - Predicted) / Observed, NA),
  ) %>%
  # Agregar columna de diagnóstico de calidad
  mutate(
    Quality = case_when(
      Abs_Residual < 0.10 ~ "Excellent",
      Abs_Residual < 0.20 ~ "Good",
      Abs_Residual < 0.30 ~ "Regular",
      TRUE ~ "Poor"
    )
  ) %>%
  # Reorganizar columnas para mejor visualización
  select(id_sample, entity_name, latitude, longitude, elevation, hill_n2,
         Observed, Predicted, Residual, Abs_Residual, Pct_Error, Squared_Error, Relative_Error, Quality,
         everything())

# Ver el resultado
head(latam_diagnostic[, 1:15])
```

Chunk 2: Crear dataframe diagnóstico para EUROPE (mejor target)
```{r}
# Extraer el mejor resultado de EUROPE
europe_best_result <- europe_results[[best_europe_target]]

# Extraer los datos preparados originales para ese target
europe_best_data <- europe_prepared[[best_europe_target]]


# Verificar que tienen el mismo número de filas
cat("Número de filas en obs_pred:", nrow(europe_best_result$obs_pred), "\n")
cat("Número de filas en datos preparados:", nrow(europe_best_data), "\n")




# Crear dataframe diagnóstico para LATAM
europe_diagnostic <- europe_best_data %>%
  # Unir con observed y predicted
  bind_cols(
    Observed = europe_best_result$obs_pred$Observed,
    Predicted = europe_best_result$obs_pred$Predicted) %>%
  # Calcular métricas de error
  mutate(
    Residual = Observed - Predicted,
    Abs_Residual = abs(Residual),
    Pct_Error = (Abs_Residual / (Observed + 0.001)) * 100,  # +0.001 para evitar división por cero
    Squared_Error = (Observed - Predicted)^2,
    Relative_Error = ifelse(Observed > 0, (Observed - Predicted) / Observed, NA),
  ) %>%
  # Agregar columna de diagnóstico de calidad
  mutate(
    Quality = case_when(
      Abs_Residual < 0.10 ~ "Excellent",
      Abs_Residual < 0.20 ~ "Good",
      Abs_Residual < 0.30 ~ "Regular",
      TRUE ~ "Poor"
    )
  ) %>%
  # Reorganizar columnas para mejor visualización
  select(id_sample, entity_name, latitude, longitude, elevation, hill_n2,
         Observed, Predicted, Residual, Abs_Residual, Pct_Error, Squared_Error, Relative_Error, Quality,
         everything())

# Ver el resultado
head(europe_diagnostic[, 1:15])
```




```{r}
# LATAM - mejores y peores predicciones
latam_best_sites <- latam_diagnostic %>%
  arrange(Abs_Residual) %>%
  select( Observed, Predicted, Abs_Residual, Quality) %>%
  head(100)

latam_worst_sites <- latam_diagnostic %>%
  arrange(desc(Abs_Residual)) %>%
  select(, Observed, Predicted, Abs_Residual, Quality) %>%
  head(100)

# EUROPE - mejores y peores predicciones
europe_best_sites <- europe_diagnostic %>%
  arrange(Abs_Residual) %>%
  select(id_sample, entity_name, latitude, longitude, elevation, Observed, Predicted, Abs_Residual, Quality) %>%
  head(200)

europe_worst_sites <- europe_diagnostic %>%
  arrange(desc(Abs_Residual)) %>%
  select(id_sample, entity_name, latitude, longitude, elevation, Observed, Predicted, Abs_Residual, Quality) %>%
  head(200)
```




```{r}
p1 <- ggplot(latam_diagnostic, aes(x = Abs_Residual)) +
  geom_histogram(bins = 30, fill = "darkgreen", alpha = 0.7) +
  geom_vline(xintercept = median(latam_diagnostic$Abs_Residual), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = paste("LATAM:", best_latam_target),
       subtitle = paste("Median =", 
                        round(median(latam_diagnostic$Abs_Residual), 4)),
       x = "Absolute error", y = "Freq") +
  theme_bw()

p1

p2 <- ggplot(europe_diagnostic, aes(x = Abs_Residual)) +
  geom_histogram(bins = 30, fill = "darkgreen", alpha = 0.7) +
  geom_vline(xintercept = median(europe_diagnostic$Abs_Residual), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = paste("EUROPE:", best_europe_target),
       subtitle = paste("Median =", 
                        round(median(europe_diagnostic$Abs_Residual), 4)),
       x = "Absolute error", y = "Freq") +
  theme_bw()

p2
```



```{r}
# Calcular rangos extendidos (15% más allá de los datos)
min_val_lat <- min(c(latam_diagnostic$Observed, latam_diagnostic$Predicted), na.rm = TRUE)
max_val_lat <- max(c(latam_diagnostic$Observed, latam_diagnostic$Predicted), na.rm = TRUE)
range_ext_lat <- (max_val_lat - min_val_lat) * 0.05
plot_min_lat <- min_val_lat - range_ext_lat
plot_max_lat <- max_val_lat + range_ext_lat
  
  
p3 <- ggplot() +
  geom_point(data = latam_diagnostic, aes(x = Observed, y = Predicted, color = Quality), alpha = 0.6, size = 2) +
   geom_point(data = latam_worst_sites, aes(x = Observed, y = Predicted),shape=8, size = 0.4, colour="gray20") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  geom_smooth(
  data = latam_diagnostic,
  aes(x = Observed, y = Predicted),
  method = "lm",
  se = FALSE,
  color = "firebrick3",
  linewidth = 0.8
) +

  coord_fixed(ratio = 1, xlim = c(plot_min_lat, plot_max_lat), ylim = c(plot_min_lat, plot_max_lat)) +
  labs(title = paste("LATAM:", best_latam_target)) +
  theme_bw()

p3



#########



# Calcular rangos extendidos (15% más allá de los datos)
min_val_eur <- min(c(europe_diagnostic$Observed, europe_diagnostic$Predicted), na.rm = TRUE)
max_val_eur <- max(c(europe_diagnostic$Observed, europe_diagnostic$Predicted), na.rm = TRUE)
range_ext_eur <- (max_val_eur - min_val_eur) * 0.05
plot_min_eur <- min_val_eur - range_ext_eur
plot_max_eur <- max_val_eur + range_ext_eur
  
  
p4 <- ggplot() +
  geom_point(data = europe_diagnostic, aes(x = Observed, y = Predicted, color = Quality), alpha = 0.6, size = 2) +
   geom_point(data = europe_worst_sites, aes(x = Observed, y = Predicted),shape=8, size = 0.4, colour="gray20") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  geom_smooth(
  data = europe_diagnostic,
  aes(x = Observed, y = Predicted),
  method = "lm",
  se = FALSE,
  color = "firebrick3",
  linewidth = 0.8
) +

  coord_fixed(ratio = 1, xlim = c(plot_min_eur, plot_max_eur), ylim = c(plot_min_eur, plot_max_eur)) +
  labs(title = paste("EUROPE:", best_europe_target)) +
  theme_bw()

p4


```



```{r}
lims <- range(c(latam_diagnostic$Observed,
                latam_diagnostic$Predicted),
              na.rm = TRUE)
breaks <- pretty(lims, n = 5)

p5 <- ggplot() +
  geom_sf(data = land_polyg_latam) +
  geom_point(data = latam_diagnostic,
             aes(x = longitude, y = latitude, colour = Observed)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "ag_Sunset", rev = TRUE),
    limits = lims,
    breaks = breaks
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = NULL) +
  theme_bw()
p5

p6 <- ggplot() +
  geom_sf(data = land_polyg_latam) +
  geom_point(data = latam_diagnostic,
             aes(x = longitude, y = latitude, colour = Predicted)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "ag_Sunset", rev = TRUE),
    limits = lims,
    breaks = breaks
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = NULL) +
  theme_bw()
p6
```


```{r}
# Modificar los plots (sin título individual)
p5 <- p5 + 
  labs(colour = "Tree fract") +
  theme(legend.text = element_text(angle = 90, hjust = 0.5, vjust = 0.5),
        legend.title = element_text(angle = 0, hjust = 0.5, vjust = 0.8))

p6 <- p6 + 
  labs(colour = "Tree fract") +
  theme(legend.text = element_text(angle = 90, hjust = 0.5, vjust = 0.5),
        legend.title = element_text(angle = 0, hjust = 0.5, vjust = 0.8))

# Crear el arreglo
fig <- ggarrange(p5, p6, 
          ncol = 2, 
          common.legend = TRUE,
          legend = "bottom",
          labels = c("A) Observed", "B) Predicted"),
          font.label = list(size = 12, face = "bold", color = "black"),
          hjust = -0.2,
          vjust = 0.6)

# Añadir título común
annotate_figure(fig,
                top = text_grob(paste("LATAM:", best_latam_target), 
                               face = "bold", 
                               size = 12))
```




Observed vs predicted for Europe

```{r}
lims <- range(c(europe_diagnostic$Observed,
                europe_diagnostic$Predicted),
              na.rm = TRUE)
breaks <- pretty(lims, n = 5)

p7 <- ggplot() +
  geom_sf(data = land_polyg_eur) +
  geom_point(data = europe_diagnostic,
             aes(x = longitude, y = latitude, colour = Observed)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "ag_Sunset", rev = TRUE),
    limits = lims,
    breaks = breaks
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = NULL) +
  theme_bw()
p7

p8 <- ggplot() +
  geom_sf(data = land_polyg_eur) +
  geom_point(data = europe_diagnostic,
             aes(x = longitude, y = latitude, colour = Predicted)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "ag_Sunset", rev = TRUE),
    limits = lims,
    breaks = breaks
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = NULL) +
  theme_bw()
p8
```


```{r}
# Modificar los plots (sin título individual)
p7 <- p7 + 
  labs(colour = "Tree fract") +
  theme(legend.text = element_text(angle = 90, hjust = 0.5, vjust = 0.5),
        legend.title = element_text(angle = 0, hjust = 0.5, vjust = 0.8))

p8 <- p8 + 
  labs(colour = "Tree fract") +
  theme(legend.text = element_text(angle = 90, hjust = 0.5, vjust = 0.5),
        legend.title = element_text(angle = 0, hjust = 0.5, vjust = 0.8))

# Crear el arreglo
fig <- ggarrange(p7, p8, 
          ncol = 2, 
          common.legend = TRUE,
          legend = "bottom",
          labels = c("A) Observed", "B) Predicted"),
          font.label = list(size = 12, face = "bold", color = "black"),
          hjust = -0.2,
          vjust = 0.6)

# Añadir título común
annotate_figure(fig,
                top = text_grob(paste("EUROPE:", best_latam_target), 
                               face = "bold", 
                               size = 12))
```








Absolute residual an squared error for LATAM

```{r}
# 1. Calcular límites comunes para Squared_Error y Abs_Residual
lims_residuals <- range(c(latam_diagnostic$Squared_Error,
                          latam_diagnostic$Abs_Residual),
                        na.rm = TRUE)
breaks_residuals <- pretty(lims_residuals, n = 5)

# 2. Re-crear los plots con escala común y sin títulos/subtítulos
p9 <- ggplot() +
  geom_sf(data = land_polyg_latam) +
  geom_point(data = latam_diagnostic,
             aes(x = longitude, y = latitude, colour = Squared_Error)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "Batlow", rev = TRUE),
    limits = lims_residuals,      # <-- Escala común
    breaks = breaks_residuals     # <-- Cortes comunes
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = NULL, colour = "Residuals") +  # Título de leyenda común
  ggtitle("Squared Error") +  # Título individual
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.text = element_text(angle = 90, hjust = 0.5, vjust = 0.5))

p10 <- ggplot() +
  geom_sf(data = land_polyg_latam) +
  geom_point(data = latam_diagnostic,
             aes(x = longitude, y = latitude, colour = Abs_Residual)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "Batlow", rev = TRUE),
    limits = lims_residuals,      # <-- Escala común
    breaks = breaks_residuals     # <-- Cortes comunes
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = NULL, colour = "Residuals") +  # Título de leyenda común
  ggtitle("Absolute Residual") +  # Título individual
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.text = element_text(angle = 90, hjust = 0.5, vjust = 0.5))

# 3. Juntar con ggarrange
library(ggpubr)

fig_residuals <- ggarrange(p9, p10, 
                           ncol = 2, 
                           common.legend = TRUE,
                           legend = "bottom",
                           labels = NULL)  # Sin labels porque ya tienen títulos

# 4. Añadir título común (opcional)
annotate_figure(fig_residuals,
                top = text_grob(paste("LATAM:", best_latam_target), 
                               face = "bold", 
                               size = 16))
```





Absolute residual an squared error for EUROPE

```{r}
# 1. Calcular límites comunes para Squared_Error y Abs_Residual
lims_residuals <- range(c(europe_diagnostic$Squared_Error,
                          europe_diagnostic$Abs_Residual),
                        na.rm = TRUE)
breaks_residuals <- pretty(lims_residuals, n = 5)

# 2. Re-crear los plots con escala común y sin títulos/subtítulos
p9 <- ggplot() +
  geom_sf(data = land_polyg_eur) +
  geom_point(data = europe_diagnostic,
             aes(x = longitude, y = latitude, colour = Squared_Error)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "Batlow", rev = TRUE),
    limits = lims_residuals,      # <-- Escala común
    breaks = breaks_residuals     # <-- Cortes comunes
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = NULL, colour = "Residuals") +  # Título de leyenda común
  ggtitle("Squared Error") +  # Título individual
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.text = element_text(angle = 90, hjust = 0.5, vjust = 0.5))

p10 <- ggplot() +
  geom_sf(data = land_polyg_eur) +
  geom_point(data = europe_diagnostic,
             aes(x = longitude, y = latitude, colour = Abs_Residual)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "Batlow", rev = TRUE),
    limits = lims_residuals,      # <-- Escala común
    breaks = breaks_residuals     # <-- Cortes comunes
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = NULL, colour = "Residuals") +  # Título de leyenda común
  ggtitle("Absolute Residual") +  # Título individual
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.text = element_text(angle = 90, hjust = 0.5, vjust = 0.5))

# 3. Juntar con ggarrange
library(ggpubr)

fig_residuals <- ggarrange(p9, p10, 
                           ncol = 2, 
                           common.legend = TRUE,
                           legend = "bottom",
                           labels = NULL)  # Sin labels porque ya tienen títulos

# 4. Añadir título común (opcional)
annotate_figure(fig_residuals,
                top = text_grob(paste("EUROPE:", best_latam_target), 
                               face = "bold", 
                               size = 16))
```


```{r}
library(patchwork)

# Escala para LATAM
lims_latam <- range(latam_diagnostic$Abs_Residual, na.rm = TRUE)
breaks_latam <- pretty(lims_latam, n = 5)

p11 <- ggplot() +
  geom_sf(data = land_polyg_latam) +
  geom_point(data = latam_diagnostic,
             aes(longitude, latitude, colour = Abs_Residual)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "Batlow", rev = TRUE),
    limits = lims_latam,
    breaks = breaks_latam,
    name = "Absolute\nResidual"
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(
    title = paste("LATAM", best_latam_target),
    x = NULL,
    y = NULL
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

# Escala para EUROPE
lims_europe <- range(europe_diagnostic$Abs_Residual, na.rm = TRUE)
breaks_europe <- pretty(lims_europe, n = 5)

p12 <- ggplot() +
  geom_sf(data = land_polyg_eur) +
  geom_point(data = europe_diagnostic,
             aes(longitude, latitude, colour = Abs_Residual)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "Batlow", rev = TRUE),
    limits = lims_europe,
    breaks = breaks_europe,
    name = "Absolute\nResidual"
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(
    title = paste("EUROPE", best_europe_target),
    x = NULL,
    y = NULL
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

# Combinar manteniendo leyendas separadas
combined_abs <- p11 + p12 +
  plot_layout(ncol = 2, guides = "keep")

combined_abs
```





```{r}
# 1. Calcular límites comunes para elevation (de ambos datasets)
lims_elev <- range(c(latam_worst_sites$elevation,
                     latam_best_sites$elevation),
                   na.rm = TRUE)
breaks_elev <- pretty(lims_elev, n = 5)

# 2. Re-crear los plots con escala correcta
p13 <- ggplot() +
  geom_sf(data = land_polyg_latam) +
  geom_point(data = latam_worst_sites, 
             aes(x = longitude, y = latitude, colour = Abs_Residual)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "Hawaii", rev = F)#,
    #limits = lims_elev,      # <-- Límites de elevación
   # breaks = breaks_elev     # <-- Cortes de elevación
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = NULL, colour = "Elevation") +  # Título de leyenda
  ggtitle("Worst 100 sites") +  # Título identificador
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))




# 1. Calcular límites comunes para elevation (de ambos datasets)
lims_elev <- range(c(europe_worst_sites$elevation,
                     europe_best_sites$elevation),
                   na.rm = TRUE)
breaks_elev <- pretty(lims_elev, n = 5)

# 2. Re-crear los plots con escala correcta
p14 <- ggplot() +
  geom_sf(data = land_polyg_eur) +
  geom_point(data = europe_worst_sites, 
             aes(x = longitude, y = latitude, colour = Abs_Residual)) +
  scale_color_gradientn(
    colours = hcl.colors(11, "Hawaii", rev = F)#,
   # limits = lims_elev,      # <-- Límites de elevación
   # breaks = breaks_elev     # <-- Cortes de elevación
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = NULL, colour = "Elevation") +  # Título de leyenda
  ggtitle("Worst 200 sites") +  # Título identificador
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))


# 3. Verificar los plots individualmente
p13
p14
```


```{r}
library(ggpubr)

# Juntar con leyenda común
fig_sites <- ggarrange(p4, p5, 
                       ncol = 2, 
                       common.legend = TRUE,
                       legend = "bottom",
                       labels = NULL)  # Sin labels porque tienen títulos

# Añadir título común
annotate_figure(fig_sites,
                top = text_grob("Elevation at Best and Worst Sites", 
                               face = "bold", 
                               size = 16))
```




```{r}
# Resumen estadístico de los errores
cat("\n=== RESUMEN DIAGNÓSTICO LATAM (", best_latam_target, ") ===\n")
latam_diagnostic %>%
  summarise(
    n = n(),
    Mean_Observed = mean(Observed),
    Mean_Predicted = mean(Predicted),
    Mean_Residual = mean(Residual),
    Mean_Abs_Residual = mean(Abs_Residual),
    Median_Abs_Residual = median(Abs_Residual),
    RMSE = sqrt(mean(Squared_Error)),
    Max_Abs_Residual = max(Abs_Residual),
    Pct_Excelente = mean(Quality == "Excelente") * 100,
    Pct_Bueno = mean(Quality == "Bueno") * 100,
    Pct_Regular = mean(Quality == "Regular") * 100,
    Pct_Pobre = mean(Quality == "Pobre") * 100
  ) %>%
  print()

cat("\n=== RESUMEN DIAGNÓSTICO EUROPE (", best_europe_target, ") ===\n")
europe_diagnostic %>%
  summarise(
    n = n(),
    Mean_Observed = mean(Observed),
    Mean_Predicted = mean(Predicted),
    Mean_Residual = mean(Residual),
    Mean_Abs_Residual = mean(Abs_Residual),
    Median_Abs_Residual = median(Abs_Residual),
    RMSE = sqrt(mean(Squared_Error)),
    Max_Abs_Residual = max(Abs_Residual),
    Pct_Excelente = mean(Quality == "Excelente") * 100,
    Pct_Bueno = mean(Quality == "Bueno") * 100,
    Pct_Regular = mean(Quality == "Regular") * 100,
    Pct_Pobre = mean(Quality == "Pobre") * 100
  ) %>%
  print()
```









#########
#########
#########


Chunk: Graficar observado vs predicho para todos los targets (LATAM)
```{r}
# Función para crear gráficos de todos los targets de LATAM
walk2(
  latam_results,
  names(latam_results),
  ~ {
    p <- ggplot(.x$obs_pred, aes(x = Observed, y = Predicted)) +
      geom_point(alpha = 0.5, size = 1.5, color = "dodgerblue2") +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
      geom_smooth(method = "lm", se = FALSE, color = "firebrick2", linewidth = 0.7) +
      coord_fixed() +
      labs(title = paste("LATAM -", .y),
           subtitle = paste("RMSEP =", round(.x$performance$RMSEP, 4),
                            " | R² =", round(.x$performance$R2, 4),
                            " | Component:", .x$best_comp)) +
      theme_bw()
    
    print(p)
  }
)
```






